---
title: "Week 2 Introduction to R"
author: "Hyun Koo"
date: "2026-01-22"
output: html_document
---

## Your first code example: Just for fun, we will try a package called coways: it is a fun package that displays text messages using ASCII art animals, making R output more playful and interactive.

```{r installing packages - Practice}

# This is where you write your code! The output will be displayed in the Console.

#install.packages("cowsay")  # installs the package from CRAN onto your computer
library(cowsay) # loads the package so you can use its functions (once you install a package, you do not need to reinstall it again. Next time, just use library().)


say("Hello world!")
say("Dr.Koo's class is so cool.\n --Anonymous",
    by = "cow")
say("I'm EGGS-hausted!", "chicken")
say("catfact", "cat")  # Displays an interesting fact using the Cat Facts API, presented in a cat-themed ASCII art.
say("It is going to snow this weekend. Stay warm!", "snowman")


# Try different animals


?say # Tip: If you type ? followed by a function name, the Help window will open on the right and explain what the function does

```

### If you enter an animal name that is not supported, R will return an error such as: Error: unknown animal. The message will also display a list of supported animals; for example frog, ghost, pumpkin, dragon, bat, owl, duck, yoda, alligator



# Practice Question and Answer
### Calculate how many years have passed since the first computer was built in 1946 and since the first moon landing in 1969. Print each result and the total number of years

```{r How to code - Practice Question and Answer}

# Historical events
first_computer <- 1946
first_moon_landing <- 1969
current_year <- 2026 #ENTER THE CURRENT YEAR
  
# Calculate years since events
years_since_computer <- current_year - first_computer
print(years_since_computer)

years_since_moon <- current_year - first_moon_landing
print(years_since_moon)

# Sum of years since events
total_years <- years_since_computer + years_since_moon
print(total_years)

```

