---
title: "1. Introduction to R"
author: "G.H. Koo"
---

In this first tutorial, we begin with the core building blocks of R. We will explore data structures in more detail in Chapter 2.

```{r}
# --------------------------------------------------
# Creating objects and basic arithmetic
# --------------------------------------------------

# Store values using the assignment operator <-
x <- 10
y <- 5

# Perform basic calculations
x + y
x * y

# --------------------------------------------------
# Creating and exploring a data frame
# --------------------------------------------------

# Create a simple data frame of cities
cities <- data.frame(
  name = c("Washington DC", "Paris", "Seoul"),
  population = c(693645, 2048472, 9600000)
)

# View the data frame
cities

# Check column names
colnames(cities)

# --------------------------------------------------
# Accessing data
# --------------------------------------------------

# Access a column using $
cities$population

# Access rows and columns using indexing [row, column]
cities[1, ]          # First row
cities[, "name"]     # Name column

# --------------------------------------------------
# Using functions
# --------------------------------------------------

# Calculate summary statistics
mean(cities$population)
sd(cities$population)

# --------------------------------------------------
# Creating new variables
# --------------------------------------------------

# Add a new column
cities$double_population <- cities$population * 2

# View updated data frame
cities
```

## Your first code example

As a warm-up, we begin with a **cowsay** package to run our initial codes.

::: {.callout-note}
The **cowsay** package displays messages using ASCII art animals.
This makes console output more fun and engaging.
:::

Install the package from CRAN onto your computer and load the package (once you install a package, you do not need to re-install it again. Next time, just use library() without install.packages().

```{r install-cowsay, message=FALSE, warning=FALSE}
#install.packages("cowsay")  
library(cowsay)
say("Hello world!")
say("I'm EGGS-hausted!.\n - Fluffy chicken",
    by = "chicken")
```

::: {.callout-tip}
Tip: If you type ? followed by a function name, the Help window will open on the right and explain what the function does
:::
```{r}
?say
```

::: {.callout-note}
Try different animals! If you enter an animal name that is not supported, R will return an error such as: Error: unknown animal. The message will also display a list of supported animals; for example frog, ghost, pumpkin, dragon, bat, owl, duck, yoda, alligator.
:::

```{r}
say("catfact", "cat")
say("It is going to snow this weekend. Stay warm! 
    — Dr. Koo", "snowman")
```

## Practice: Years Since Major Historical Events

::: {.callout-tip}
**Task**

1. Calculate how many years have passed since: the first computer (1946) + the first moon landing (1969).

2. Print each result and the total number of years.
:::

```{r How to code - Practice Question and Answer}
first_computer <- 1946
first_moon_landing <- 1969
current_year <- 2026 #ENTER THE CURRENT YEAR
```

#### Calculate years since events
```{r}
years_since_computer <- current_year - first_computer
print(years_since_computer)

years_since_moon <- current_year - first_moon_landing
print(years_since_moon)
```

#### Sum of years since events
```{r}
total_years <- years_since_computer + years_since_moon
print(total_years)
```

::: {.callout-note} 
**Tip: Remember to label your R code chunks**

When writing code in your R Markdown, make sure that every R code block starts with three backticks and {r}, and also ends with three backticks. The backtick key is usually located next to the 1 key on your keyboard (the same key as ~).

:::