---
title: "1. Introduction to R"
author: "Hyun Koo"
---

## Your first code example

As a warm-up, we begin with a **cowsay** package to run our initial codes.

::: {.callout-note}
The **cowsay** package displays messages using ASCII art animals.
This makes console output more fun and engaging.
:::

```{r install-cowsay, message=FALSE, warning=FALSE}
#install.packages("cowsay")  # installs the package from CRAN onto your computer
library(cowsay) # loads the package so you can use its functions (once you install a package, you do not need to reinstall it again. Next time, just use library().)
say("Hello world!")
say("We are so cool.\n --Anonymous",
    by = "cow")
say("I'm EGGS-hausted!", "chicken")
say("catfact", "cat")
say("It is going to snow this weekend. Stay warm!", "snowman")

?say
```

::: {.callout-tip}
Tip: If you type ? followed by a function name, the Help window will open on the right and explain what the function does
:::

::: {.callout-note}
Try different animals! If you enter an animal name that is not supported, R will return an error such as: Error: unknown animal. The message will also display a list of supported animals; for example frog, ghost, pumpkin, dragon, bat, owl, duck, yoda, alligator.
:::

## Practice: Years Since Major Historical Events

::: {.callout-tip}
**Task**
Calculate how many years have passed since: the first computer (1946) + the first moon landing (1969)
Print each result and the total number of years.
:::

```{r How to code - Practice Question and Answer}
first_computer <- 1946
first_moon_landing <- 1969
current_year <- 2026 #ENTER THE CURRENT YEAR
  
# Calculate years since events
years_since_computer <- current_year - first_computer
print(years_since_computer)

years_since_moon <- current_year - first_moon_landing
print(years_since_moon)

# Sum of years since events
total_years <- years_since_computer + years_since_moon
print(total_years)

```

