---
title: "8. Network analysis"
author: "G.H. Koo"
---

## Learning goals

By the end of this tutorial, you will be able to:

- Understand the difference between **semantic networks** and **social networks**.
- Build a **semantic network** from bigrams in text data.
- Build a **social network** from user mentions and replies.
- Interpret common network measures such as **degree**, **betweenness**, and **community structure**.
- Visualize networks using **igraph** and **ggraph**.

---

# Introduction to Network Analysis

Network analysis helps us study relationships rather than isolated observations. In text analysis, this can mean examining how words co-occur with one another. In social media research, it can mean examining how users interact through replies and mentions.

In this tutorial, we cover two related applications:

- **Semantic network analysis**, which maps connections among words
- **Social network analysis**, which maps connections among users

::: {.callout-note}
A network has two basic parts:

- **nodes**, which represent words or users
- **edges**, which represent relationships between them
:::

---

## Required Packages

You only need to install packages once.

```{r eval=FALSE}
packages <- c(
  "tidyverse",
  "tidytext",
  "tokenizers",
  "igraph",
  "ggraph",
  "tidygraph",
  "stringr"
)
install.packages(packages)
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(tokenizers)
library(tidytext)
library(igraph)
library(ggraph)
library(tidygraph)
library(stringr)
set.seed(381)
options(scipen = 999)
```

## Importing data

For this tutorial, we use a CSV file containing abortion-related Twitter data.

```{r}
tweets <- read.csv("~/Desktop/abortion_tweets.csv", header = TRUE, sep = ",")

tweets_subset <- tweets %>%
  select(
    id,
    author_id,
    created_at,
    text,
    public_metrics.impression_count,
    public_metrics.like_count,
    public_metrics.quote_count,
    public_metrics.reply_count,
    public_metrics.retweet_count,
    referenced_tweets,
    in_reply_to_user_id,
    entities.mentions
  )

head(tweets_subset)
```

---

# Part 1: Semantic Network Analysis

## What is a semantic network?

A semantic network shows how words are connected based on their co-occurrence in text. Here, we use **bigrams** (two-word combinations) to identify words that frequently appear together.

## Step 1: Clean retweet prefixes

We begin by removing leading `RT @username` patterns so they do not dominate the network.

```{r}
tweets_cleaned <- tweets_subset %>%
  mutate(text = str_remove_all(text, "^RT @\\w+"))
```

## Step 2: Extract bigrams and remove stopwords

```{r}
tw_bigram <- tweets_cleaned %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  separate(bigram, into = c("word1", "word2"), sep = " ") %>%
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word
  ) %>%
  unite(bigram, word1, word2, sep = " ") %>%
  count(bigram, sort = TRUE)

head(tw_bigram)
```

## Step 3: Separate bigrams into two words

```{r}
bigram_separated <- tw_bigram %>%
  separate(bigram, into = c("word1", "word2"), sep = " ")
```

## Step 4: Filter infrequent bigrams

```{r}
bigram_filtered <- bigram_separated %>%
  filter(n > 100)
```

::: {.callout-tip}
The threshold used here (`n > 100`) is just an example. Lower thresholds keep more word pairs but can make the network harder to read.
:::

## Step 5: Build the network object

```{r}
bigram_graph <- graph_from_data_frame(bigram_filtered)
```

## Step 6: Visualize the semantic network

```{r}
ggraph(bigram_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = name), repel = TRUE) +
  labs(title = "Semantic Network of Bigrams in Abortion Tweets") +
  theme_void()
```

## Optional: Add directional edges

```{r}
direction_edges <- grid::arrow(type = "closed", length = unit(.2, "cm"))

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), arrow = direction_edges) +
  geom_node_point(size = 2) +
  geom_node_text(aes(label = name), repel = TRUE) +
  labs(title = "Semantic Network of Bigrams with Directional Information") +
  theme_void()
```

## Describe the semantic network

```{r}
vcount(bigram_graph)
ecount(bigram_graph)
edge_density(bigram_graph)
components(bigram_graph)$no
```

## Questions for interpretation

1. Which words appear most connected?
2. Are there clusters of words representing different themes?
3. Do certain words act as bridges between themes?
4. What does this suggest about how abortion is discussed in the dataset?

---

# Part 2: Social Network Analysis

In social network analysis, users are the nodes and interactions among users are the edges.

## Step 1: Prepare interaction data

```{r}
tweets_interaction <- tweets_subset %>%
  select(author_id, in_reply_to_user_id, entities.mentions)
```

## Step 2: Build reply and mention edge lists

```{r}
replies <- tweets_interaction %>%
  filter(!is.na(in_reply_to_user_id)) %>%
  select(source = author_id, target = in_reply_to_user_id)

mentions <- tweets_interaction %>%
  filter(!is.na(entities.mentions)) %>%
  unnest(entities.mentions) %>%
  select(source = author_id, target = entities.mentions)

replies <- replies %>%
  mutate(source = as.character(source), target = as.character(target))

mentions <- mentions %>%
  mutate(source = as.character(source), target = as.character(target))

edges <- bind_rows(replies, mentions) %>%
  distinct()
```

## Step 3: Filter to highly connected users

```{r}
user_counts <- edges %>%
  pivot_longer(cols = c(source, target), values_to = "user") %>%
  count(user) %>%
  filter(n > 500)

user_counts
```

::: {.callout-note}
A high count means a user appears often in the interaction data, but it does not automatically mean that the user is substantively important.
:::

## Step 4: Build and visualize the user network

```{r}
filtered_edges <- edges %>%
  filter(source %in% user_counts$user, target %in% user_counts$user)

user_graph <- graph_from_data_frame(filtered_edges, directed = TRUE)

interaction_graph <- ggraph(user_graph, layout = "fr") +
  geom_edge_link(aes(alpha = 0.5), arrow = arrow(type = "closed", length = unit(0.15, "cm"))) +
  geom_node_point() +
  geom_node_text(aes(label = name), repel = TRUE) +
  labs(title = "Twitter User Interaction Network") +
  theme_void()

interaction_graph
```

## Interpret user interactions

In this graph:

- nodes represent Twitter users
- edges represent mentions or replies
- more connected nodes indicate users with more ties to others in the network

If you want to inspect a specific user ID, you can sometimes search the platform directly using a format like `from:USER_ID`, as long as the account is still public and available.

## Step 5: Degree and betweenness centrality

```{r}
degree_centrality <- degree(user_graph, mode = "all")
top_users <- sort(degree_centrality, decreasing = TRUE)[1:10]
print(top_users)

betweenness_centrality <- betweenness(user_graph, directed = TRUE)
top_influencers <- sort(betweenness_centrality, decreasing = TRUE)[1:10]
print(top_influencers)
```

## Extra: Basic network description

```{r}
vcount(user_graph)
ecount(user_graph)
edge_density(user_graph)
components(user_graph)$no
```

### How to interpret these numbers

- More nodes means more unique users are involved.
- More edges means more interactions are taking place.
- Higher density means users are more interconnected.
- More components means the network contains separate groups that are not connected to each other.

## Extra: Compare in-degree and out-degree

```{r}
in_degree <- degree(user_graph, mode = "in")
top_in_degree <- sort(in_degree, decreasing = TRUE)[1:10]
print(top_in_degree)

out_degree <- degree(user_graph, mode = "out")
top_out_degree <- sort(out_degree, decreasing = TRUE)[1:10]
print(top_out_degree)
```

### How to interpret this

- High in-degree users are frequently mentioned or replied to.
- High out-degree users are highly active in reaching out to others.
- These are not always the same users.

## Extra: Detect communities in the network

```{r}
user_graph_undirected <- as_undirected(user_graph, mode = "collapse")
comm <- cluster_louvain(user_graph_undirected)
length(comm)
modularity(comm)
```

## Extra: Visualize communities

```{r}
community_graph <- as_tbl_graph(user_graph_undirected) %>%
  activate(nodes) %>%
  mutate(community = as.factor(membership(comm)))

community_plot <- ggraph(community_graph, layout = "fr") +
  geom_edge_link(alpha = 0.15) +
  geom_node_point(aes(color = community), size = 3) +
  labs(title = "Communities in the Twitter Interaction Network") +
  theme_void()

print(community_plot)
```

---

# Summary

In this tutorial, you learned how to:

- construct a semantic network from bigrams
- visualize relationships among words using `igraph` and `ggraph`
- construct a social network from user mentions and replies
- interpret common network measures such as degree, betweenness, density, and modularity

Network analysis is useful because it helps us move from isolated units of text or users to the broader relational structure that shapes communication.
