---
title: "7. Sentiment analysis"
author: "G.H. Koo"
---

## Learning goals

By the end of this tutorial, you will be able to:

- Understand what **dictionary-based sentiment analysis** is and when it is useful.
- Preprocess social media text for sentiment analysis in R.
- Apply the **AFINN lexicon** to estimate tweet-level sentiment.
- Summarize and interpret sentiment scores across tweets.
- Explore whether sentiment is associated with **engagement metrics**.

---

# Introduction to Sentiment Analysis

Sentiment analysis is a computational method used to identify the **emotional tone** of text. In communication research, it is often used to classify text as positive, negative, or neutral, or to estimate emotional intensity using predefined dictionaries.

In this tutorial, we use a **dictionary-based approach** to analyze sentiment in Twitter data containing the keyword **"abortion"** from **June 24, 2022, to December 31, 2022**.

Because social media text is unstructured, the first step is always **cleaning and preprocessing** the data. This helps reduce noise and improves the clarity of the analysis.

::: {.callout-note}
Since this dataset was collected when the platform was still called **Twitter**, I refer to it as Twitter rather than X in this tutorial.
:::

::: {.callout-tip}
As of February 2023, Twitter (now X) no longer provides free API access. If you need social media data for practice, you may want to explore public archives such as:

- [Harvard Dataverse](https://dataverse.harvard.edu/)
- [r/datasets](https://www.reddit.com/r/datasets/)
- [Twitter/X API documentation](https://developer.x.com/en/docs/x-api)
:::

---

## Required Packages

You only need to install packages once.

```{r eval=FALSE}
packages <- c(
  "tidyverse",
  "tidytext",
  "tokenizers",
  "stringr"
)
install.packages(packages)
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(tokenizers)
library(tidytext)
library(stringr)
set.seed(381)
options(scipen = 999)
```

## Importing data

For this tutorial, we use a CSV file containing abortion-related Twitter data.

```{r, warning=FALSE, message=FALSE}
tweets <- read.csv("~/Desktop/abortion_tweets.csv", header = TRUE, sep = ",")

tweets_subset <- tweets %>%
  select(
    id,
    author_id,
    created_at,
    text,
    public_metrics.impression_count,
    public_metrics.like_count,
    public_metrics.quote_count,
    public_metrics.reply_count,
    public_metrics.retweet_count,
    referenced_tweets
  )
```

---

# Part 1: Preprocessing Text

## Step 1: Tokenize the text

To prepare the tweets for sentiment analysis, we:

- preserve the original text
- remove URLs
- tokenize the tweets into individual words
- convert all words to lowercase
- remove stopwords

```{r}
tweets_tokenized <- tweets_subset %>%
  mutate(original_text = text) %>%
  mutate(text = str_remove_all(text, "\\bhttps[^\\s]+")) %>%
  unnest_tokens(word, text, to_lower = TRUE) %>%
  anti_join(stop_words, by = "word")

head(tweets_tokenized)
```

::: {.callout-tip}
If you want to inspect the most common words before running sentiment analysis, try:

```{r eval=FALSE}
tweets_tokenized %>%
  count(word, sort = TRUE)
```
:::

---

# Part 2: Dictionary-Based Sentiment Analysis

## What is the AFINN lexicon?

In dictionary-based sentiment analysis, we use a predefined lexicon that assigns sentiment values to words.

In this tutorial, we use the AFINN lexicon, which assigns words scores ranging from -5 (very negative) to +5 (very positive).

Other commonly used lexicons include:

- Bing: positive / negative
- NRC: eight emotions plus positive / negative
- Lexicoder: often used in political communication
- LIWC: sentiment and broader psychological categories

Each dictionary has different strengths and limitations. Because language is complex, researchers should always interpret dictionary results carefully.

```{r}
afinn_dictionary <- get_sentiments("afinn")
head(afinn_dictionary)
subset(afinn_dictionary, word == "love")
subset(afinn_dictionary, word == "terrible")
```

## Step 2: Match tweet words to the sentiment dictionary

We use `inner_join()` to keep only the words in our dataset that also appear in the AFINN dictionary.

```{r}
tweets_sentiment <- tweets_tokenized %>%
  inner_join(afinn_dictionary, by = "word")

head(tweets_sentiment)
```

The `value` column shows the sentiment score assigned to each word.

## Step 3: Identify the most negative and most positive words

```{r}
most_negative_words <- tweets_sentiment %>%
  arrange(value) %>%
  select(word, value) %>%
  distinct() %>%
  head(10)
most_negative_words
```

```{r}
most_positive_words <- tweets_sentiment %>%
  arrange(desc(value)) %>%
  select(word, value) %>%
  distinct() %>%
  head(10)
most_positive_words
```

These outputs help us see which words in the dataset contribute most strongly to negative and positive sentiment.

## Step 4: Calculate tweet-level sentiment scores

Next, we summarize the sentiment values within each tweet. Since `id` is the unique tweet identifier, we group by `id` and sum the sentiment values.

```{r}
tweets_summarized <- tweets_sentiment %>%
  group_by(id) %>%
  summarize(sentiment = sum(value), .groups = "drop")
head(tweets_summarized)
```

Now we merge these sentiment scores back into the original tweet-level dataset.

```{r}
tweets_final <- tweets_subset %>%
  full_join(tweets_summarized, by = "id")
```

## Step 5: Replace missing sentiment values with 0

Some tweets will not contain any words found in the AFINN dictionary. In those cases, sentiment will appear as `NA`. We replace those missing values with `0`.

```{r}
tweets_final$sentiment[is.na(tweets_final$sentiment)] <- 0
```

---

# Part 3: Describing and Interpreting Sentiment

## Distribution of sentiment scores

```{r}
table(tweets_final$sentiment)
```

This gives a simple overview of how sentiment is distributed across the tweets.

## Inspect negative and positive tweets

Negative tweets:

```{r}
negative_tweets <- tweets_final %>%
  filter(sentiment <= -1) %>%
  select(text, sentiment)
head(negative_tweets)
```

Positive tweets:

```{r}
positive_tweets <- tweets_final %>%
  filter(sentiment >= 1) %>%
  select(text, sentiment)
head(positive_tweets)
```

Looking directly at tweet examples helps researchers evaluate whether the sentiment scores make sense in context.

::: {.callout-note}
This step is important because dictionary-based sentiment analysis cannot fully capture sarcasm, irony, humor, or context-dependent meanings.
:::

---

# Part 4: Sentiment and Engagement

## Step 1: Examine the relationship with likes

As a simple extension, we can test whether more positive or negative tweets tend to receive more likes.

```{r}
cor.test(
  tweets_final$sentiment,
  tweets_final$public_metrics.like_count,
  method = "spearman",
  use = "complete.obs"
)
```

## Step 2: Visualize sentiment and likes

```{r}
ggplot(tweets_final, aes(x = sentiment, y = public_metrics.like_count)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(
    title = "Sentiment and Likes in Abortion Tweets",
    x = "Sentiment score",
    y = "Like count"
  ) +
  theme_minimal()
```

::: {.callout-tip}
Engagement variables such as likes, replies, and retweets are often highly skewed. In substantive research, it is a good idea to inspect their distributions before choosing a model.
:::

---

# Summary

In this tutorial, you learned how to:

- preprocess tweets for dictionary-based sentiment analysis
- apply the AFINN lexicon to estimate tweet-level sentiment
- identify strongly positive and negative words in the dataset
- interpret sentiment scores using both summary tables and tweet examples
- explore whether sentiment is associated with a simple engagement metric

Dictionary-based sentiment analysis is easy to implement and useful for introductory text analysis, but it should always be interpreted with care because sentiment depends heavily on context.
